---
title: "Annotate the Plot"
author: "Zuguang Gu (z.gu@dkfz.de)"
date: '`r Sys.Date()`'
output: 
  rmarkdown::html_vignette:
    fig_caption: true
    css: main.css
vignette: >
  %\VignetteIndexEntry{Annotate the Plot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 5,
    fig.height = 5,
    fig.align = "center"
)
options(width = 100)
```

In this vignette, I demonstrate how to add text and legend to the spiral plot.


```{r, echo = FALSE}
library(spiralize)
library(ComplexHeatmap)
library(circlize)
```


## Text

**spiralize** is implemented under **grid** graphics system, thus functions
from **grid** need to be used with **spiralize**.

The spiral plot is drawn in a square viewport, as shown in the following figure,
thus, simply you can use `grid.text()` to add text to a specific position.

```{r}
library(spiralize)
spiral_initialize()
spiral_track()
grid.rect(gp = gpar(fill = NA))
grid.text("center", 0, 0, default.units = "native")
grid.text("top_left", 0, 1, default.units = "npc", just = c("left", "top"))
```


For time series data, users might want to add texts to describe different "seasons/sectors". In the following example,
assume period is day, I add hour names to the corresponding sectors in the polar coordinate system. I also add text
to the start and end of the spiral.

```{r}
spiral_initialize_by_time(xlim = c("2021-01-01 00:00:00", "2021-01-04 23:59:59"), padding = unit(1.5, "cm"))
spiral_track()

s = current_spiral()

d = seq(360/24/2, 360, by = 360/24) %% 360
for(i in seq_along(d)) {
    df = polar_to_cartesian(d[i]/180*pi, (s$max_radius + 1.5))
    grid.text(paste0(i, "h"), x = df[1, 1], y = df[1, 2], default.unit = "native",
        rot = ifelse(d[i] > 0 & d[i] < 180, d[i] - 90, d[i] + 90), gp = gpar(fontsize = 10))
}

x = s$get_numeric_x(c("2021-01-01 00:00:00", "2021-01-04 23:59:59"))
df = xy_to_cartesian(x, 0.5)
grid.text("2012-01-01\n00:00:00", df$x[1], df$y[1], default.units = "native", vjust = 1.2,
    gp = gpar(fontsize = 8))
grid.text("2012-04-01\n23:59:59", df$x[2], df$y[2], default.units = "native", vjust = -0.2,
    gp = gpar(fontsize = 8))
```

## Legend

Legends are important graphics elements in a plot which guild users to understand the plot.
**spiralize** does not directly draw legends, but it can be very easily done by other legend functions,
such as `Legend()` and `packLegend()` from **ComplexHeatmap** package. 

`Legend()` creates a single legend, either continuous or discrete, and `packLegend()` combines a list
of legends into a single object. Check the following example:

```{r, fig.width = 7}
n = 1000
spiral_initialize(xlim = c(0, n))
spiral_track(height = 0.4)

x = runif(n)
col_fun = circlize::colorRamp2(c(0, 0.5, 1), c("green", "white", "red"))
spiral_rect(1:n - 1, 0, 1:n, 1, gp = gpar(fill = col_fun(x), col = NA))

spiral_highlight(10, 100, type = "line", gp = gpar(col = "blue"))
spiral_highlight(500, 700, type = "line", gp = gpar(col = "purple"))

library(ComplexHeatmap)
lgd = packLegend(
    Legend(title = "title1", col_fun = col_fun),
    Legend(title = "title2", type = "lines", legend_gp = gpar(col = c("blue", "purple"), lwd = 2),
        at = c("text1", "text2"))
)
draw(lgd, x = unit(1, "npc") + unit(2, "mm"), just = "left")
```

When I draw the legend, I put the legend to a position of `unit(1, "npc") +
unit(2, "mm")` which is 2mm to the right of the spiral plot. To make the
legend properly visible, you must set the image width larger than the image
height in order to give space to the legend (in the image above, the width is
7 inch and the height is 5 inch).


If you also want to add title for the plot, you need to do a little bit more. One simple
way is to push several viewports, as shown in the folowing code. 


```{r, fig.width = 7}
grid.newpage()
title_height = grobHeight(textGrob("A", gp = gpar(fontsize = 14)))*2
pushViewport(viewport(y = 1, height = title_height, just = "top"))
grid.text("This is a title", gp = gpar(fontsize = 14))
popViewport()

pushViewport(viewport(y = 0, height = unit(1, "npc") - title_height, just = "bottom"))
spiral_initialize(xlim = c(0, n), newpage = FALSE)  # here setting newpage = FALSE is important!
spiral_track(height = 0.4)
spiral_rect(1:n - 1, 0, 1:n, 1, gp = gpar(fill = col_fun(x), col = NA))

spiral_highlight(10, 100, type = "line", gp = gpar(col = "blue"))
spiral_highlight(500, 700, type = "line", gp = gpar(col = "purple"))

draw(lgd, x = unit(1, "npc") + unit(2, "mm"), just = "left")

spiral_clear()
popViewport()
```

One thing that needs to be noted here is, since here we are manipulating viewports, 
it is suggested to call `spiral_clear()` when the spiral plot is done so that the arrangement
of viewports won't be messed up.


There is also a convinient way to construct legend for horizon chart. Function `spiral_horizon()`
actually returns an object which later can be used in `horizon_legend()` to generate a legend where
the legend labels are properly formatted.

The following plot shows the downloads of **ggplot2** over time:

```{r, fig.width = 8, results = "hide"}
df = readRDS(system.file("extdata", "ggplot2_downloads.rds", package = "spiralize"))
day_diff = as.double(df$date[nrow(df)] - df$date[1], "days")
year_mean = tapply(df$count, lubridate::year(df$date), function(x) mean(x[x > 0]))

df$diff = log2(df$count/year_mean[as.character(lubridate::year(df$date))])
df$diff[is.infinite(df$diff)] = 0
q = quantile(abs(df$diff), 0.99)
df$diff[df$diff > q] = q
df$diff[df$diff < -q] = -q

spiral_initialize_by_time(xlim = range(df$date))
spiral_track(height = 0.9)
lt = spiral_horizon(df$date, df$diff, use_bars = TRUE)

lgd = horizon_legend(lt, title = "log2(difference)")
draw(lgd, x = unit(1, "npc") + unit(2, "mm"), just = "left")
```