---
title: "Initialize Spirals by Special Data Types"
author: "Zuguang Gu (z.gu@dkfz.de)"
date: '`r Sys.Date()`'
output: 
  rmarkdown::html_vignette:
    fig_caption: true
---

```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    warning = FALSE,
    fig.width = 7,
    fig.height = 7,
    fig.align = "center",
    fig.retina = 2
)
knitr::knit_hooks$set(pngquant = knitr::hook_pngquant)
knitr::opts_chunk$set(
  message = FALSE,
  dev = "ragg_png",
  fig.align = "center",
  pngquant = "--speed=10 --quality=30"
)
options(width = 100)
```

```{r, echo = FALSE}
library(spiralize)
library(cowplot)
library(GetoptLong)
```


## Work with date/time object

For time series data, values on _x_-axis are time points in a certain unit, e.g.
days, or hours, and are linearly distributed along the axis. To map time
series data to the spiral, internally a simple conversion is applied. Assume
the first time point is `t1` and the last time point is `t2`, the number of time points is `n = t2
- t1` which is the time difference between the two. If we take `[0, 1]` for the
first time point, `[1, 2]` for the second time point, and `[n-1, n]` for the last time point, then
internally the spiral is initialized with `xlim = c(0, n)` and a time point `d` is
converted to the internal numeric value by `d - d1 + 0.5`.

The unit of time point can be set via the argument `unit_on_axis` and the unit
of the period can be set via the argument `period`, e.g. `unit_on_axis =
"days"` and `period = "years"`. `unit_on_aixs` can be set as one of `"days"`,
`"months"`, `"weeks"`, `"hours"`, `"mins"` and `"secs"`. And there are also
corresponding values for `period`. If these two arguments are not set,
they are guessed from `xlim` automatically.

In the following examples, also note the default value of `polar_lines_by` is also different for different `period`. 
E.g. there are 12 polar lines for years (12 months), 7 polar lines for weeks (7 weekdays), 24 polar lines for days (24 hours).

```{r}
spiral_initialize_by_time(xlim = c("2014-01-01", "2021-06-17"))
spiral_track(height = 0.6)
spiral_axis()

spiral_initialize_by_time(xlim = c("2021-01-01 00:00:00", "2021-01-05 00:00:00"))
spiral_track(height = 0.6)
spiral_axis()

spiral_initialize_by_time(xlim = c("2021-01-01 00:00:00", "2021-01-01 00:10:00"),
    unit_on_axis = "secs", period = "mins")
spiral_track(height = 0.6)
spiral_axis()
```


As shown in previous examples, the values for `xlim` should be a time/date
object or their character forms that can be converted to the time/date object.
Later when adding graphics, the time/date objects or their character forms can
also be used as the _x_-locations for the low-level graphics functions. For example:

```{r, eval = FALSE}
spiral_points("2021-01-01", 0.5)
```

Argument `at` can be set to a vector of times to control the breaks on the axis.
In the next example, I set the time interval to every two hours.

```{r, message = FALSE}
spiral_initialize_by_time(xlim = c("2021-01-01 00:00:00", "2021-01-05 00:00:00"))
spiral_track(height = 0.6)

library(lubridate)
# `by` is measured in seconds
at = seq(as.POSIXlt("2021-01-01 00:00:00"), as.POSIXlt("2021-01-05 00:00:00"), by = 2*60*60)
spiral_axis(at = at)
```

In the last example of this section, I let the spiral go clockwisely and put 0/12 o'clock on the 
top of each loop (simply done by setting `clockwise = TRUE`) and each loop corresponds to 12 hours:


```{r}
spiral_initialize_by_time(xlim = c("2021-01-01 00:00:00", "2021-01-03 00:00:00"),
    start = 360 + 90, period_per_loop = 0.5, clockwise = TRUE)
spiral_track(height = 0.6)
at = seq(as.POSIXlt("2021-01-01 00:00:00"), as.POSIXlt("2021-01-05 00:00:00"), by = 2*60*60)
spiral_axis(at = at)
```

When each loop corresponds to a year and units on axis are days, maybe you
have already seen the messages in the previous examples, by default, each loop
actually corresponds to only 52 weeks (364 days) and the remaining 1 or 2
days will be added and accumulated to the next year. This might be a problem
when there are many loops (many years). In the following left plot where we
visualize 10 years, the total accumuated years result in about 15 degrees exceeding
the full circle. The function `spiral_initialize_by_time()` allows to set an
argument `normalize_year = TRUE` so that each loop is perfectly a year. But
users need to be cautious that now it is not easy to correspond weekdays
between years because they are not perfectly aligned.


```{r, eval = FALSE}
# the left plot
spiral_initialize_by_time(xlim = c("2010-01-01", "2019-12-31"))
spiral_track()
for(i in 0:9) {
    # years() is from lubridate package
    spiral_text(as.POSIXlt("2010-01-01") + years(i), 0.5, 2010 + i, facing = "inside")
}

# the right plot
spiral_initialize_by_time(xlim = c("2010-01-01", "2019-12-31"),
    normalize_year = TRUE)
spiral_track()
for(i in 0:9) {
    spiral_text(as.POSIXlt("2010-01-01") + years(i), 0.5, 2010 + i, facing = "inside")
}

```

```{r, echo = FALSE, fig.width = 10, fig.height = 5, message = FALSE}
p1 = grid.grabExpr({
    spiral_initialize_by_time(xlim = c("2010-01-01", "2019-12-31"))
    spiral_track()
    for(i in 0:9) {
        spiral_text(as.POSIXlt("2010-01-01") + years(i), 0.5, 2010 + i, facing = "inside")
    }
    grid.text("default", 0, 1, just = c("left", "top"), gp = gpar(fontsize = 14))
}, width = 5, height = 5)
p2 = grid.grabExpr({
    spiral_initialize_by_time(xlim = c("2010-01-01", "2019-12-31"),
        normalize_year = TRUE)
    spiral_track()
    for(i in 0:9) {
        spiral_text(as.POSIXlt("2010-01-01") + years(i), 0.5, 2010 + i, facing = "inside")
    }
    grid.text("normalize_year = TRUE", 0, 1, just = c("left", "top"), gp = gpar(fontsize = 14))
}, width = 5, height = 5)
plot_grid(p1, p2)
```

Last, when each loop corresponds to a year and units on axis are days, and
when `normalize_year` is set to `TRUE`, the background sectors partitioned by
dashed radical lines actually correspond to the proportion of month days in a year, i.e., the
width of Feburary is the smallest.

```{r}
spiral_initialize_by_time(xlim = c("2010-01-01", "2014-12-31"), normalize_year = TRUE)
spiral_track()
for(i in 0:11) {
    d = as.POSIXlt("2014-01-15") + months(i)
    spiral_text(d, 1.5, month.name[month(d)], gp = gpar(fontsize = 8), 
        facing = "inside", nice_facing = TRUE)
}
```

## Work with genomic coordinates

Genomic data can be represented as a data frame (e.g. in _bed_ format), or as
a `GRanges` object. To initialize the spiral for genomic data visualization,
the data on _x_-axis corresponds to the genomic coordinates. The spiral with
genomic data can be initialized by the function
`spiral_initialize_by_gcoor()`.  The genomic coordinates are linear numeric
values, thus, `spiral_initialize_by_gcoor()` is basically the same as
`spiral_initialize()`. The only difference is the axis labels are
automatically formatted for genomic coordinates.

Also since normally there are no periodic patterns for genomc data, `scale_by`
is by default set to `"curve_length"`.

```{r}
spiral_initialize_by_gcoor(xlim = c(2e6, 8e6))  # 2MB to 8MB
spiral_track(height = 0.6)
spiral_axis()
spiral_points(x = runif(500, min = 2e6, max = 8e6), runif(500), pch = 16, gp = gpar(col = 2))
```
