---
title: "Initialize Spirals by Special Data Types"
author: "Zuguang Gu (z.gu@dkfz.de)"
date: '`r Sys.Date()`'
output: 
  rmarkdown::html_vignette:
    fig_caption: true
    css: main.css
vignette: >
  %\VignetteIndexEntry{2. Initialize Spirals by Special Data Types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 7,
    fig.height = 7,
    fig.align = "center"
)
options(width = 100)
```

```{r, echo = FALSE}
library(spiralize)
library(cowplot)
library(GetoptLong)
```

## Work with date/time object

For time series data, values on _x_-axis are time points in a certain unit, e.g.
days, or hours, and are linearly distributed along the axis. To map time
series data to the spiral, internally a simple conversion is applied. Assume
the first time point is `t1` and the last time point is `t2`, the number of time points is `n = t2
- t1` which is the time difference between the two. If we take `[0, 1]` for the
first time point, `[1, 2]` for the second time point, and `[n-1, n]` for the last time point, then
internally the spiral is initialized with `xlim = c(0, n)` and a time point `d` is
converted to the internal numeric value by `d - d1 + 0.5`.

The unit of time point can be set via the argument `unit_on_axis` and the unit
of the period can be set via the argument `period`, e.g. `unit_on_axis =
"days"` and `period = "years"`. `unit_on_aixs` can be set as one of `"days"`,
`"months"`, `"weeks"`, `"hours"`, `"mins"` and `"secs"`. And there are also
corresponding valid values for `period`. If these two arguments are not set,
they are guessed from `xlim` automatically.

In the following examples, also note the default value of `polar_lines_by` is also different for different `period`. 
E.g. there are 12 polar lines for years (12 months), 7 polar lines for weeks (7 weekdays), 24 polar lines for days (24 hours).

```{r}
spiral_initialize_by_time(xlim = c("2014-01-01", "2021-06-17"))
spiral_track(height = 0.6)
spiral_axis()

spiral_initialize_by_time(xlim = c("2021-01-01 00:00:00", "2021-01-05 00:00:00"))
spiral_track(height = 0.6)
spiral_axis()

spiral_initialize_by_time(xlim = c("2021-01-01 00:00:00", "2021-01-01 00:10:00"),
    unit_on_axis = "secs", period = "mins")
spiral_track(height = 0.6)
spiral_axis()
```


As shown in previous examples, the values for `xlim` should be a time/date object or an object that can be converted to the time/date object.
Later when adding graphics, the time/date objects can also be used as the _x_-locations for the low-level graphics functions. e.g.

```{r, eval = FALSE}
spiral_points("2021-01-01", 0.5)
```

## Work with genomic coordinates

Genomic data can be represented as a data frame (e.g. in _bed_ format), or as
a `GRanges` object. To initialize the spiral for genomic data visualization,
the data on _x_-axis corresponds to the genomic coordinates. The spiral with
genomic data can be initialized by the function
`spiral_initialize_by_gcoor()`.  The genomic coordinates are linear numeric
values, thus, `spiral_initialize_by_gcoor()` is basically the same as
`spiral_initialize()`. The only difference is the axis labels are
automatically formated for genomic coordinates.

Also since normally there are no periodic patterns for genomc data, `scale_by`
is by default set to `"curve_length"`.

```{r}
spiral_initialize_by_gcoor(xlim = c(2e6, 8e6))  # 2MB to 8MB
spiral_track(height = 0.6)
spiral_axis()
spiral_points(x = runif(500, min = 2e6, max = 8e6), runif(500), pch = 16, gp = gpar(col = 2))
```
