---
title: "Visualize Data on Spirals"
author: "Zuguang Gu (z.gu@dkfz.de)"
date: '`r Sys.Date()`'
output: 
  rmarkdown::html_vignette:
    fig_caption: true
    css: main.css
    toc: true
vignette: >
  %\VignetteIndexEntry{Visualize Data on Spirals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 5,
    fig.height = 5,
    fig.align = "center"
)
options(width = 100)
```

```{r, echo = FALSE}
library(spiralize)
library(cowplot)
library(GetoptLong)
```

In this vignette, I describe the package **spiralize** which visualizes data along an Archimedean spiral.
It has two major advantages for visualization:

1. It is able to visualize data with very long axis with high resolution.
2. It is efficient for time series data to reveal periodic patterns.


## The Archimedean spiral

In polar coordinates ($r$, $\theta$), the [Archimedean
spiral](https://en.wikipedia.org/wiki/Archimedean_spiral) has the following form:

$$ r = b \cdot \theta $$

where $b$ controls the distance between two loops. The radical distance between two neighbouring loops for a given $\theta$ is:

$$ d(\theta) = r(\theta + 2\pi) - r(\theta) = b \cdot (\theta + 2\pi) - b \cdot \theta = b \cdot 2\pi $$

This shows the radical distance between two neighbouring loops is independent to the value of $\theta$ and is a constant value. The following 
figure demonstrates an Archimedean spiral with 4 loops ($\theta \in [0, 8\pi]$).


```{r, echo = FALSE}
theta = seq(0, 360*4, by = 1)/180*pi
b = 1/2/pi
r = theta*b

df = spiralize:::polar_to_cartesian(theta, r)

abs_max_x = max(abs(df$x))
abs_max_y = max(abs(df$y))

grid.newpage()
  
padding = unit(c(10, 10), "mm")
pushViewport(viewport(
    width = unit(1, "snpc") - padding[1], 
    height = unit(1, "snpc") - padding[2],
    xscale = c(-abs_max_x, abs_max_x), 
    yscale = c(-abs_max_y, abs_max_y)))

d = seq(0, 360, by = 30)
if(d[length(d)] == 360) d = d[-length(d)]
dm = matrix(nrow = length(d), ncol = 4)
for(i in seq_along(d)) {
  r0 = max(r + b*2*pi)*1.1
  dm[i, ] = c(0, 0, cos(d[i]/180*pi)*r0, sin(d[i]/180*pi)*r0)
}
grid.segments(dm[, 1], dm[, 2], dm[, 3], dm[, 4], default.units = "native", 
    gp = gpar(col = "grey", lty = 3))

grid.lines(df$x, df$y, default.units = "native")

grid.segments(2, 0, 3, 0, arrow = arrow(angle = 20, length = unit(2, "mm"), ends = "both"), default.units = "native")
grid.text("d", unit(2.5, "native"), unit(0, "native") + unit(1, "mm"), just = "bottom")
popViewport()
```

Note $\theta$ can also be negative values where the spiral is mirrored by _y_-axis
(in Cartesian coordinates). In **spiralize**, we only consider $\theta$ as
positive values. The mirrored spiral can be set by the `flip` argument which is introduced later
in this vignette.

Since the distance between any two neighbouring loops for any given $\theta$ is
constant, it is a ideal place to put tracks along the spiral where the tracks
have identical radical heights everywhere. Later the tracks can be served as
virtual coordinate systems to map to data. This is why the package is called
"_spiralize_" (to transform a normal Cartesian coordinate system to a curved
spiral shape). The following two figures demonstrate a spiral with one track
and with two tracks. The red line is the spiral itself. The spiral ranges
between $\pi/2$ and $6\pi$. It is easy to see the upper border of each track
is also a spiral but with an offset $a$:


$$ r = a + b \cdot \theta $$

where $a$ is the offset to the "Base spiral" (the red spiral in the following plots).

```{r, fig.width = 10, fig.height = 5, echo = FALSE}
grid.newpage()
pushViewport(viewport(x = 0.25, width = 0.5))
spiral_initialize(start = 90, end = 360*3, newpage = FALSE)
spiral_track(height = 0.45)
spiral_lines(TRACK_META$xlim, rep(get_track_data("ymin"), 2), gp = gpar(col = "red"))
spiral_text(2/11, 0.5, "track 1")
spiral_clear()
popViewport()

pushViewport(viewport(x = 0.75, width = 0.5))
spiral_initialize(start = 90, end = 360*3, newpage = FALSE)
spiral_track(height = 0.45)
spiral_lines(TRACK_META$xlim, rep(get_track_data("ymin"), 2), gp = gpar(col = "red"))
spiral_text(2/11, 0.5, "track 1")
spiral_track(height = 0.45, background_gp = gpar(fill = "#CCCCCC"))
spiral_text(2/11, 0.5, "track 2")
spiral_clear()
popViewport()
```

Denote the maximal radius of the spiral as $d_{max} = b \cdot \theta_{max}$,
and denote the length of the spiral as $l$ (which has a complex form, see https://downloads.imagej.net/fiji/snapshots/arc_length.pdf), we can
treat $d_{max}$ as the resolution of the visualization applied in the normal
Cartesian coordinate system and $l$ as the resolution of the visualization
applied on the spiral. Then the ratio of the two resolutions is:

$$ ratio = \frac{l}{2d_{max}} $$

E.g., for a spiral with 5 loops ($\theta_{max} = 10\pi$), the ratio is 7.89,
which means the spiral improves the resolution of visualization almost to 8
folds. Generally, the ratio increases almost linearly to the number of loops.

```{r, echo = FALSE, fig.keep = "none"}
spiral_initialize()
s = current_spiral()
```

```{r, echo = FALSE}
ratio = function(k) {
    theta = pi*2*k
    b = 1/2/pi
    r = b*theta
    len = s$spiral_length(theta)
    len/2/r
}

k = 1:20
plot(k, ratio(k), xlab = "number of loops in the spiral", ylab = "ratio", 
    main = "ratio of the resolution between\nspiral and normal Cartesian system")
```

The relationship between ratio and $\theta$ has the following form:

$$ ratio = \frac{\mathrm{ln}(\theta + \sqrt{1 + \theta^2})}{4\theta} + \frac{\sqrt{1 + \theta^2}}{4} $$


## The layout of the spiral

The function `spiral_initialize()` is used to intialize the spiral. Arguments `start` and `end` control
the angular range of the spiral. Here the values should be in degrees and they are converted to radians internally.

In **spiralize**, the parameter $b$ in the spiral equation $r = b \cdot \theta$ is set to $b = 1/2\pi$, so that the distance
between two neighbouring loops is $d = 1$. Denote $\theta_e$ as the
end angle (in radians) of the spiral, the ranges of the viewport (under **grid** graphics system) on both _x_-axis and _y_-axis that draw the spiral are $[-x, x]$ where

$$x = b \cdot \theta_e + d = 1/2\pi \cdot \theta_e + 1$$


The following two plots demonstrate different values of `start` and `end`. Also as shown in the following example code,
I suggest to set the values of `start` and `end` in a form of `360*a + b`, e.g. `360*4 + 180`, so that it is straighforward
to know the positions in the polar coordinates and how many loops there are in the spiral (I think people should feel more natural with degrees than radians).

```{r, eval = FALSE}
# the left plot
spiral_initialize(start = 90, end = 360)
spiral_track()

# the right plot
spiral_initialize(start = 180, end = 360*4 + 180)
spiral_track()
```

```{r, fig.width = 10, fig.height = 5, echo = FALSE}
grid.newpage()
pushViewport(viewport(x = 0.25, width = 0.5))

spiral_initialize(start = 90, end = 360, newpage = FALSE)
spiral_track()
spiral_clear()
popViewport()

pushViewport(viewport(x = 0.75, width = 0.5))

spiral_initialize(start = 180, end = 360*4 + 180, newpage = FALSE)
spiral_track()
spiral_clear()
popViewport()
```

Argument `flip` controls how to flip the spiral. It accpets one of the four values: `"none"`/`"horizontal"`/`"vertical"`/`"both"`.
Examples are as follows. In the examples, I additionally add the axes in the tracks to show in which direction the data extends along
the spiral. I also manually adjust the height of the track to give enough space for axes.

```{r, eval = FALSE}
# the top left plot
spiral_initialize(flip = "none") # default
spiral_track(height = 0.6)
spiral_axis()

# the top right plot
spiral_initialize(flip = "horizontal")
spiral_track(height = 0.6)
spiral_axis()

# the bottom left plot
spiral_initialize(flip = "vertical")
spiral_track(height = 0.6)
spiral_axis()

# the bottom right plot
spiral_initialize(flip = "both")
spiral_track(height = 0.6)
spiral_axis()
```

```{r, fig.width = 10, fig.height = 10, echo = FALSE}
p1 = grid.grabExpr({
    spiral_initialize()
    spiral_track(height = 0.6)
    spiral_axis()
    grid.text("flip = 'none'", 0, 1, just = c("left", "top"), gp = gpar(fontsize = 14))
})
p2 = grid.grabExpr({
    spiral_initialize(flip = "horizontal")
    spiral_track(height = 0.6)
    spiral_axis()
    grid.text("flip = 'horizontal'", 0, 1, just = c("left", "top"), gp = gpar(fontsize = 14))
})
p3 = grid.grabExpr({
    spiral_initialize(flip = "vertical")
    spiral_track(height = 0.6)
    spiral_axis()
    grid.text("flip = 'vertical'", 0, 1, just = c("left", "top"), gp = gpar(fontsize = 14))
})
p4 = grid.grabExpr({
    spiral_initialize(flip = "both")
    spiral_track(height = 0.6)
    spiral_axis()
    grid.text("flip = 'both'", 0, 1, just = c("left", "top"), gp = gpar(fontsize = 14))
})
plot_grid(p1, p2, p3, p4, nrow = 2)
```

Argument `scale_by` controls how to linearly scale the data on the
spiral. It allows value of `"angle"` or `"curve_length"` (or for short,
`"curve"`). `"angle"` means equal difference on data corresponds to equal
difference of angles in the polar coordinates. `"curve_length"` means equal
difference on data corresponds to equal difference of the length of the
spiral. Observe how the axis ticks distribute in the following two plots. Also
the polar lines are removed for `scale_by = "curve_length"`.

```{r, eval = FALSE}
# the left plot
spiral_initialize(scale_by = "angle") # default
spiral_track(height = 0.6)
spiral_axis()

# the right plot
spiral_initialize(scale_by = "curve_length")
spiral_track(height = 0.6)
spiral_axis()
```

```{r, fig.width = 10, fig.height = 5, echo = FALSE}
p1 = grid.grabExpr({
    spiral_initialize(scale_by = "angle")
    spiral_track(height = 0.6)
    spiral_axis()
    grid.text("scale_by = 'angle'", 0, 1, just = c("left", "top"), gp = gpar(fontsize = 14))
})
p2 = grid.grabExpr({
    spiral_initialize(scale_by = "curve_length")
    spiral_track(height = 0.6)
    spiral_axis()
    grid.text("scale_by = 'curve_length'", 0, 1, just = c("left", "top"), gp = gpar(fontsize = 14))
})
plot_grid(p1, p2)
```

The following heatmaps might be clearer to demonstrate the difference between `"angle"` and `"curve_length"`. In this 
example, each grid has the equal bin size of the data.


```{r, fig.width = 10, fig.height = 5, echo = FALSE}
make_plot = function(scale_by) {
    n = 100
    col = circlize::colorRamp2(c(0, 0.5, 1), c("blue", "white", "red"))
    spiral_initialize(xlim = c(0, n), scale_by = scale_by)
    spiral_track(height = 0.9)

    x = runif(n)
    spiral_rect(1:n - 1, 0, 1:n, 1, gp = gpar(fill = col(x), col = NA))
    grid.text(qq("scale_by = '@{scale_by}'"), 0, 1, just = c("left", "top"), gp = gpar(fontsize = 14))
}
p1 = grid.grabExpr(make_plot("angle"))
p2 = grid.grabExpr(make_plot("curve_length"))
plot_grid(p1, p2)
```

As you can see, when `scale_by` is set to `"angle"`, in outer loops, even when the actually difference
on data is the same, the physical widths are larger than these in inner loops. Nevertheless, when the data is 
time series or periodic, `"angle"` is the proper choice because it is easy to directly compare between loops
which are the same time points over different periods. As a comparison, `"curve_length"` won't provide any
periodic information.

The spiral grows from inner loops to outer loops, thus, by default, data increases from
the inner loops as well. This can be reversed by setting argument `reverse = TRUE`. See
the following example and also observe the axes. The red arrows indicate the direction of axes.

```{r, eval = FALSE}
# the left plot
spiral_initialize(reverse = FALSE) # default
spiral_track()
spiral_arrow(0.2, 0.8, gp = gpar(fill = "red"))
spiral_axis()

# the right plot
spiral_initialize(reverse = TRUE)
spiral_track()
spiral_arrow(0.2, 0.8, gp = gpar(fill = "red"))
spiral_axis()
```

```{r, fig.width = 10, fig.height = 5, echo = FALSE}
p1 = grid.grabExpr({
    spiral_initialize(reverse = FALSE)
    spiral_track()
    spiral_arrow(0.2, 0.8, gp = gpar(fill = "red"))
    spiral_axis()
    grid.text("reverse = FALSE", 0, 1, just = c("left", "top"), gp = gpar(fontsize = 14))
})
p2 = grid.grabExpr({
    spiral_initialize(reverse = TRUE)
    spiral_track()
    spiral_arrow(0.2, 0.8, gp = gpar(fill = "red"))
    spiral_axis()
    grid.text("reverse = TRUE", 0, 1, just = c("left", "top"), gp = gpar(fontsize = 14))
})
plot_grid(p1, p2)
```

To map data to spiral, argument `xlim` should be set which corresponds to data range on _x_-axis.
Observe the axes in the following plots.

```{r, eval = FALSE}
# the left plot
spiral_initialize(xlim = c(0, 1000))
spiral_track(height = 0.6)
spiral_axis()

# the right plot
spiral_initialize(xlim = c(-1000, 1000))
spiral_track(height = 0.6)
spiral_axis()
```

```{r, echo = FALSE, fig.width = 10, fig.height = 5}
p1 = grid.grabExpr({
    spiral_initialize(xlim = c(0, 1000))
    spiral_track(height = 0.6)
    spiral_axis()
})
p2 = grid.grabExpr({
    spiral_initialize(xlim = c(-1000, 1000))
    spiral_track(height = 0.6)
    spiral_axis()
})
plot_grid(p1, p2)
```

## Create tracks

After the spiral is intialized, next we can add tracks along it. Argument `height` controls
the height of the track. The value of `height` is a value between 0 and 1 which is the fraction of the distance
between two neighbouring loops in the spiral. In the following left plot, I add black border to the track 
by setting the argument `background_gp`.

```{r, eval = FALSE}
# the left plot
spiral_initialize()
spiral_track(height = 1, background_gp = gpar(col = "black"))

# the right plot
spiral_initialize()
spiral_track(height = 0.5)
```

```{r, echo = FALSE, fig.width = 10, fig.height = 5}
p1 = grid.grabExpr({
    spiral_initialize()
    spiral_track(height = 1, background_gp = gpar(col = "black"))
    grid.text("height = 1", 0, 1, just = c("left", "top"), gp = gpar(fontsize = 14))
})
p2 = grid.grabExpr({
    spiral_initialize()
    spiral_track(height = 0.5)
    grid.text("height = 0.5", 0, 1, just = c("left", "top"), gp = gpar(fontsize = 14))
})
plot_grid(p1, p2)
```

Multiple tracks can be added sequentially. Just note the sum of heights of all tracks
should not exceed 1.

```{r}
spiral_initialize()
spiral_track(height = 0.4, background_gp = gpar(fill = 2))
spiral_track(height = 0.2, background_gp = gpar(fill = 3))
spiral_track(height = 0.1, background_gp = gpar(fill = 4))
```

Data range on _y_-axis is specified by the argument `ylim`. In the following example, I also add a _y_-axis
by `spiral_yaxis()`.

```{r}
spiral_initialize()
spiral_track(ylim = c(0, 100))
spiral_yaxis()
```

## Graphics functions

Tracks are created with data ranges on both _x_-axis and _y_-axis. Now the tracks can be thought as
normal Cartesian coordinates. There are following low-level graphics functions so that complex plots
can be easily constructed by combining these low-level graphics functions.

### Points

Like other graphics functions e.g. `points()` or `grid.points()`, the "spiral graphics functions" also
accept locations on _x_-axis and _y_-axis for data points. `spiral_points()` draws points
in the spiral track.

```{r}
spiral_initialize() # by default xlim = c(0, 1)
spiral_track()      # by default ylim = c(0, 1)
spiral_points(x = runif(1000), y = runif(1000))
```

### Lines

Adding lines with `spiral_lines()` is also straightforward:

```{r}
x = sort(runif(1000))
y = runif(1000)
spiral_initialize()
spiral_track()
spiral_lines(x, y)
```

Argument `type` can be set to `"h"` so that vertical lines (or radical lines
if you take polar coordinates as reference) are drawn to the baseline for each
data point.

```{r, eval = FALSE}
# the left plot
spiral_initialize()
spiral_track()
spiral_lines(x, y, type = "h")

# the right plot
spiral_initialize()
spiral_track()
spiral_lines(x, y, type = "h", baseline = 0.5, gp = gpar(col = ifelse(y > 0.5, "red", "blue")))
```

```{r, echo = FALSE, fig.width = 10, fig.height = 5}
p1 = grid.grabExpr({
    spiral_initialize()
    spiral_track()
    spiral_lines(x, y, type = "h")
})
p2 = grid.grabExpr({
    spiral_initialize()
    spiral_track()
    spiral_lines(x, y, type = "h", baseline = 0.5, gp = gpar(col = ifelse(y > 0.5, "red", "blue")))
})
plot_grid(p1, p2)
```

Argument `area` can be set to `TRUE` so that area under the lines can be filled with a certain color.

```{r}
spiral_initialize()
spiral_track()
spiral_lines(x, y, area = TRUE, gp = gpar(fill = 2, col = NA))
```

Note you can also set `baseline` with `area = TRUE`, however, you cannot set different colors for the area
above the baseline and below the baseline. Consider to use `spiral_bars()` or `spiral_horizon()` for this scenario.

### Segments

`spiral_segments()` draws a list of segments.


```{r}
n = 1000
x0 = runif(n)
y0 = runif(n)
x1 = x0 + runif(n, min = -0.01, max = 0.01)
y1 = 1 - y0

spiral_initialize(xlim = range(c(x0, x1)))
spiral_track()
spiral_segments(x0, y0, x1, y1, 
    gp = gpar(col = circlize::rand_color(n, luminosity = "bright"), lwd = runif(n, 0.5, 3)))
```

The same as `grid.segments()`, you can also set the argument `arrow` to add arrows on the segments.

```{r}
n = 100
x0 = runif(n)
y0 = runif(n)
x1 = x0 + runif(n, min = -0.01, max = 0.01)
y1 = 1 - y0

spiral_initialize(xlim = range(c(x0, x1)))
spiral_track()
spiral_segments(x0, y0, x1, y1, arrow = arrow(length = unit(2, "mm")),
    gp = gpar(col = circlize::rand_color(n, luminosity = "bright"), lwd = runif(n, 0.5, 3)))
```


### Rectangles

`spiral_rect()` draws rectangles, which is the base function for drawing heatmaps and barplots. The first four
arguments are the coordinates of the bottom left and top right of the rectangles.

```{r}
n = 1000
require(circlize)
spiral_initialize(xlim = c(0, n))
spiral_track(height = 0.9)

x1 = runif(n)
col1 = circlize::colorRamp2(c(0, 0.5, 1), c("blue", "white", "red"))
spiral_rect(1:n - 1, 0, 1:n, 0.5, gp = gpar(fill = col1(x1), col = NA))

x2 = runif(n)
col2 = circlize::colorRamp2(c(0, 0.5, 1), c("green", "white", "red"))
spiral_rect(1:n - 1, 0.5, 1:n, 1, gp = gpar(fill = col2(x2), col = NA))
```

### Bars

`spiral_bars()` can draw bars simply from a numeric vector. Bars can also be drawn to a baseline.

```{r, eval = FALSE}
x = seq(1, 1000, by = 1) - 0.5  # middle points of bars
y = runif(1000)

# the left plot
spiral_initialize(xlim = c(0, 1000))
spiral_track(height = 0.8)
spiral_bars(x, y)

# the right plot
spiral_initialize(xlim = c(0, 1000))
spiral_track(height = 0.8)
spiral_bars(x, y, baseline = 0.5, gp = gpar(fill = ifelse(y > 0.5, 2, 3), col = NA))
```

```{r, echo = FALSE, fig.width = 10, fig.height = 5}
x = seq(1, 1000, by = 1) - 0.5  # middle points of bars
y = runif(1000)

p1 = grid.grabExpr({
    spiral_initialize(xlim = c(0, 1000))
    spiral_track(height = 0.8)
    spiral_bars(x, y)
})

p2 = grid.grabExpr({
    spiral_initialize(xlim = c(0, 1000))
    spiral_track(height = 0.8)
    spiral_bars(x, y, baseline = 0.5, gp = gpar(fill = ifelse(y > 0.5, 2, 3), col = NA))
})
plot_grid(p1, p2)
```

`spiral_bars()` can also draw bars from a matrix, then each column in the matrix correspond to one stack of the bars.

```{r}
y = matrix(runif(3*1000), ncol = 3)
y = y/rowSums(y)
spiral_initialize(xlim = c(0, 1000))
spiral_track(height = 0.8)
spiral_bars(x, y, gp = gpar(fill = 2:4, col = NA))
```

Width of bars can be different. You can set a vector to the argument of `bar_width`. Note `x` always corresponds
to the middle of each bar.

```{r}
w = runif(100)
w = w/sum(w)  # width of bars, sum of all width is 1
b = c(0, cumsum(w))
x = (b[1:100] + b[2:101])/2  # middle of each bar
y = runif(100)

spiral_initialize()
spiral_track()
spiral_bars(x, y, bar_width = w)
```

### Polygons

`spiral_polygon()` draws polygons. Note the polygon must be closed, which means, the last data point should overlap to the first one.

```{r}
x0 = sort(runif(200))
x0 = matrix(x0, ncol = 2, byrow = TRUE)
x1 = sort(runif(200))
x1 = matrix(x1, ncol = 2, byrow = TRUE)

spiral_initialize()
spiral_track()

for(i in 1:100) {
    pt1 = circlize:::get_bezier_points(x0[i, 1], 0, x1[i, 1], 1, xlim = c(0, 1), ylim = c(0, 1))
    pt2 = circlize:::get_bezier_points(x0[i, 2], 0, x1[i, 2], 1, xlim = c(0, 1), ylim = c(0, 1))

    spiral_polygon(
        c(x0[i, 1], x0[i, 2], pt2[, 1], rev(pt1[, 1]), x0[i, 1]),
        c(0, 0, pt2[, 2], rev(pt1[, 2]), 0), 
        gp = gpar(fill = rand_color(1, luminosity = "bright"), col = NA)
    )
}
```

### Text

`spiral_text()` draws texts. Argument `facing` controls the rotation of texts.

```{r, eval = FALSE}
x = seq(0.1, 0.9, length = 26)
text = strrep(letters, 6)

# the top left plot
spiral_initialize()
spiral_track()
spiral_text(x, 0.5, text, facing = "downward") # default

# the bottom left plot
spiral_initialize()
spiral_track()
spiral_text(x, 0.5, text, facing = "inside")

# the bottom right plot
spiral_initialize()
spiral_track()
spiral_text(x, 0.5, text, facing = "outside")
```

```{r, echo = FALSE, fig.width = 10, fig.height = 10}
x = seq(0.1, 0.9, length = 26)
text = strrep(letters, 6)

p1 = grid.grabExpr({
    spiral_initialize()
    spiral_track()
    spiral_text(x, 0.5, text, facing = "downward")
    grid.text("facing = 'downward'", 0, 1, just = c("left", "top"), gp = gpar(fontsize = 14))
})

p2 = grid.grabExpr({
    grid.newpage()
})

p3 = grid.grabExpr({
    spiral_initialize()
    spiral_track()
    spiral_text(x, 0.5, text, facing = "inside")
    grid.text("facing = 'inside'", 0, 1, just = c("left", "top"), gp = gpar(fontsize = 14))
})

p4 = grid.grabExpr({
    spiral_initialize()
    spiral_track()
    spiral_text(x, 0.5, text, facing = "outside")
    grid.text("facing = 'outside'", 0, 1, just = c("left", "top"), gp = gpar(fontsize = 14))
})
plot_grid(plot_grid(p1, p2), plot_grid(p3, p4), nrow = 2)
```

Text can also be set to `"clockwise"` or `"reverse_clockwise"`:

```{r, eval = FALSE}
x = seq(0.1, 0.9, length = 26)

# the left plot
spiral_initialize()
spiral_track()
spiral_text(x, 0.5, "aaaa", facing = "clockwise")

# the right plot
spiral_initialize()
spiral_track()
spiral_text(x, 0.5, "aaaa", facing = "reverse_clockwise")
```

```{r, echo = FALSE, fig.width = 10, fig.height = 5}
x = seq(0.1, 0.9, length = 26)

p1 = grid.grabExpr({
    spiral_initialize()
    spiral_track()
    spiral_text(x, 0.5, "aaaa", facing = "clockwise")
    grid.text("facing = 'clockwise'", 0, 1, just = c("left", "top"), gp = gpar(fontsize = 14))
})

p2 = grid.grabExpr({
    spiral_initialize()
    spiral_track()
    spiral_text(x, 0.5, "aaaa", facing = "reverse_clockwise")
    grid.text("facing = 'reverse_clockwise'", 0, 1, just = c("left", "top"), gp = gpar(fontsize = 14))
})
plot_grid(p1, p2)
```

For long texts, `facing` can be set to `"curved_inside"` or `"curved_outside"` so that curved
texts are draw along the spiral.

```{r, eval = FALSE}
x = seq(0.1, 0.9, length = 10)
text = rep(paste(letters, collapse = ""), 10)

# the left plot
spiral_initialize()
spiral_track()
spiral_text(x, 0.5, text, facing = "curved_inside")

# the right plot
spiral_initialize()
spiral_track()
spiral_text(x, 0.5, text, facing = "curved_outside")
```

```{r, echo = FALSE, fig.width = 10, fig.height = 5}
x = seq(0.1, 0.9, length = 10)
text = rep(paste(letters, collapse = ""), 10)

p1 = grid.grabExpr({
    spiral_initialize()
    spiral_track()
    spiral_text(x, 0.5, text, facing = "curved_inside")
    grid.text("facing = 'curved_inside'", 0, 1, just = c("left", "top"), gp = gpar(fontsize = 14))
}, width = 5, height = 5)
p2 = grid.grabExpr({
    spiral_initialize()
    spiral_track()
    spiral_text(x, 0.5, text, facing = "curved_outside")
    grid.text("facing = 'curved_outside'", 0, 1, just = c("left", "top"), gp = gpar(fontsize = 14))
}, width = 5, height = 5)
plot_grid(p1, p2)
```

Calculation of positions of letters of the curved text depends on the size of current graphics device. When the device
changes its size, the positions of letters will not be correct and you need to regenerate the plot. Also
users need to be careful when using `grid.grabExpr()` to capture the plot. By default `grid.grabExpr()` captures
graphics output in a device with 7inch x 7inch. Users might need to manually set the device size to make sure
the curved texts are not affected.

In the next example, I use `grid.grabExpr()` to capture two spiral plots with curved texts. Later the two plots are merged
with using the **cowplot** package and the final merged plot is saved in a PDF with 10 inches width and 5 inches height. 
I manually set the device size in the two `grid.grabExpr()` calls so that the size of the place where the graphics are captured
is the same as the size of the place where they are finally drawn.

```{r, eval = FALSE}
p1 = grid.grabExpr({
    spiral_initialize()
    spiral_track()
    spiral_text(x, 0.5, text, facing = "curved_inside")
}, width = 5, height = 5)
p2 = grid.grabExpr({
    spiral_initialize()
    spiral_track()
    spiral_text(x, 0.5, text, facing = "curved_outside")
}, width = 5, height = 5)

pdf(..., width = 10, height = 5)
plot_grid(p1, p2)
dev.off()
```

one last thing for drawing text is that the argument `nice_facing` can be set to `TRUE` so that the rotation
of texts are automatically adjusted so that they are easy to read, i.e. all the texts always face the lower part of the polar coordinate system.

### Axis

`spiral_aixs()` draws axis along the spiral. So it is the _x_-axis of the data.

```{r}
spiral_initialize()
spiral_track(height = 0.6)
spiral_axis()
```

Argument `major_at` or simply `at` controls the break points on the axis and argument `labels`
controls the corresponding axis labels.

```{r, eval = FALSE}
# the left plot
spiral_initialize(xlim = c(0, 360*4), start = 360, end = 360*5)
spiral_track(height = 0.6)
spiral_axis(major_at = seq(0, 360*4, by = 30))

# the right plot
spiral_initialize(xlim = c(0, 12*4), start = 360, end = 360*5)
spiral_track(height = 0.6)
spiral_axis(major_at = seq(0, 12*4, by = 1), labels = c("", rep(month.name, 4)))
```

```{r, echo = FALSE, fig.width = 10, fig.height = 5}
p1 = grid.grabExpr({
    spiral_initialize(xlim = c(0, 360*4), start = 360, end = 360*5)
    spiral_track(height = 0.6)
    spiral_axis(major_at = seq(0, 360*4, by = 30))
})
p2 = grid.grabExpr({
    spiral_initialize(xlim = c(0, 12*4), start = 360, end = 360*5)
    spiral_track(height = 0.6)
    spiral_axis(major_at = seq(0, 12*4, by = 1), labels = c("", rep(month.name, 4)))
})
plot_grid(p1, p2)
```

If the axis labels are too long, argument `curved_labels` can be set to `TRUE` so that the labels
are curved along the spiral.

```{r}
spiral_initialize()
spiral_track(height = 0.6)
spiral_axis(at = c(0.1, 0.3, 0.6, 0.9), labels = strrep(letters[1:4], 20), curved_labels = TRUE)
```


`spiral_yaxis()` draws _y_-axis. Argument `side` controls which side of the track to put
the _y_-axis. `side` can be set to `"both"` so that _y_-axis is drawn on the two sides of the track.

```{r, eval = FALSE}
# the left plot
spiral_initialize()
spiral_track(height = 0.8)
spiral_yaxis(side = "start")
spiral_yaxis(side = "end", at = c(0, 0.25, 0.5, 0.75, 1), labels = letters[1:5])

# the right plot
spiral_initialize()
spiral_track(height = 0.8)
spiral_yaxis(side = "both")
```

```{r, echo = FALSE, fig.width = 10, fig.height = 5}
p1 = grid.grabExpr({
    spiral_initialize()
    spiral_track(height = 0.8)
    spiral_yaxis(side = "start")
    spiral_yaxis(side = "end", at = c(0, 0.25, 0.5, 0.75, 1), labels = letters[1:5])
})
p2 = grid.grabExpr({
    spiral_initialize()
    spiral_track(height = 0.8)
    spiral_yaxis(side = "both")
})
plot_grid(p1, p2)
```

### Horizon chart

Tracks along the spiral are long, but the heights of the tracks are normally
very small. [Horizon
chart](https://flowingdata.com/2015/07/02/changing-price-of-food-items-and-horizon-graphs/)
is an effcient way to visualize distributions by vertically folding the
distribution graphics, which makes it possible to visualize in a plotting
region with a very small height.

In the next example, I visualize difference of **ggplot2** daily downloads to
the mean of the current year, between 2015-01-01 to 2020-12-31.

```{r}
# The data can be downloaded by the next line. The data object is already saved in spiralize package.
# df = cranlogs::cran_downloads("ggplot2", from = "2015-01-01")
df = readRDS(system.file("extdata", "ggplot2_downloads.rds", package = "spiralize"))
# to simplify the data, I only take the complete years between 2015 and 2020
df = df[df$date < as.Date("2021-01-01"), ]
day_diff = as.double(df$date[nrow(df)] - df$date[1], "days")
year_mean = tapply(df$count, lubridate::year(df$date), function(x) mean(x[x > 0]))

df$diff = log2(df$count/year_mean[as.character(lubridate::year(df$date))])
df$diff[is.infinite(df$diff)] = 0
q = quantile(abs(df$diff), 0.99)  # adjust outliers
df$diff[df$diff > q] = q
df$diff[df$diff < -q] = -q
head(df)
```

Function `spiral_horizon()` draws the horizon chart along the spiral. The input variables are _x_-locations
and _y_-locations of the data. To align weeks at different years, each loop contains 364 (52 weeks). In the following
plot, red areas correspond to those days when daily downloads are higher than the yearly average and blue areas
correspond to the days when daily downloads are less than the yearly average.

```{r, fig.width = 6, fig.height = 6}
spiral_initialize(xlim = c(0, nrow(df)), start = 360, end = 360*(day_diff/364) + 360) # a circle of 52 weeks
spiral_track(height = 0.9)
spiral_horizon(1:nrow(df) - 0.5, df$diff)
```

Bars can be used to put on each days by setting argument `use_bar = TRUE`:

```{r, fig.width = 6, fig.height = 6}
spiral_initialize(xlim = c(0, nrow(df)), start = 360, end = 360*(day_diff/364) + 360)
spiral_track(height = 0.9)
spiral_horizon(1:nrow(df) - 0.5, df$diff, use_bars = TRUE)
```

When using bars, argument `bar_width` can be set as a vector if the bar widths are not all equal.

### Images

`spiral_raster()` adds images to the spiral. Currently it supports formats of png/svg/pdf/eps/jpeg/jpg/tiff. The formats
can be mixed used.

```{r}
image = system.file("extdata", "Rlogo.png", package = "circlize")
x = seq(0.1, 0.9, length = 10)

spiral_initialize()
spiral_track()
spiral_raster(x, 0.5, image)
```

Similar as text, argument `facing` can be set to control the rotation of images. If `nice_facing = TRUE`, the image
with rotation facing the top will be automatically adjusted to bottom.

```{r, eval = FALSE}
# the left plot
spiral_initialize()
spiral_track()
spiral_raster(x, 0.5, image, facing = "inside")

# the right plot
spiral_initialize()
spiral_track()
spiral_raster(x, 0.5, image, facing = "inside", nice_facing = TRUE)
```

```{r, echo = FALSE, fig.width = 10, fig.height = 5}
p1 = grid.grabExpr({
    spiral_initialize()
    spiral_track()
    spiral_raster(x, 0.5, image, facing = "inside")
}, width = 5, height = 5)
p2 = grid.grabExpr({
    spiral_initialize()
    spiral_track()
    spiral_raster(x, 0.5, image, facing = "inside", nice_facing = TRUE)
}, width = 5, height = 5)
plot_grid(p1, p2)
```

### Arrows

`spiral_arrow()` draws arrows along the spiral.

```{r, fig.width = 6, fig.height = 6}
spiral_initialize()
spiral_track()
spiral_arrow(0.3, 0.6, gp = gpar(fill = "red", col = NA))
spiral_arrow(0.8, 0.9, gp = gpar(fill = "blue"), tail = "point", arrow_position = "start")
```

### Highlight 

`spiral_highlight()` highlights a specific section of the spiral. If the argument `type` is set to `"rect"` (the default), it
highlights the section with a semi-transparent rectangle which covers the complete range on _y_-axis. If `type` is
set to `"line"`, an annotation line is drawn either at the bottom or on the top of the track.

```{r}
spiral_initialize()
spiral_track()
spiral_highlight(0.4, 0.6)
spiral_highlight(0.1, 0.2, type = "line", gp = gpar(col = "blue"))
spiral_highlight(0.7, 0.8, type = "line", line_side = "outside")
```

If `scale_by` is set to `angle` (the default), users might also want to
highlight a specific of interval but across several cycles, e.g. from January
to March in year 2000 to 2010. Here the function
`spiral_highlight_by_sector()` draws a semi-transparent sector to highlight a
fixed interval over several cycles.

The first four argument `x1`, `x2`, `x3` and `x4` in
`spiral_highlight_by_sector()` determine the position of the sector. If only
`x1` and `x2` are specified, the start circle is calculated from `x1` and `x2`
and the end circle is the most outside one. If `x3` and `x4` are also
specified, the outer circle is calcualted from `x3` and `x4`.

```{r}
spiral_initialize(xlim = c(0, 360*4), start = 360, end = 360*5)
spiral_track(height = 0.6)
spiral_axis()
spiral_highlight_by_sector(36, 72)
spiral_highlight_by_sector(648, 684)
spiral_highlight_by_sector(216, 252, 936, 972, gp = gpar(fill = "blue"))
```


### Dendrograms and phylogenetic trees

When the dendrograms or phlogenetic trees have huge number of leaves, it is also suitable to use spiral to visualize them. 
There are two functions: `spiral_dendrogram()` for `dendrogram` objects and `spiral_phylo()` for
`phylo` objects.


Note since dendrograms or phylogenetic trees do not have "periodic patterns",
`scale_by` in `spiral_initialize()` is usually set to `"curve_length"`.

The `dendrogram` object can be rendered with **dendextend** package. See the following examples.


```{r, eval = FALSE}
dend = as.dendrogram(hclust(dist(runif(1000))))

# the left plot
spiral_initialize(xlim = c(0, 1000), start = 360, end = 360*2 + 180, scale_by = "curve_length")
spiral_track()
spiral_dendrogram(dend)

# the right plot
library(dendextend)
dend = color_branches(dend, k = 4)
spiral_initialize(xlim = c(0, 1000), start = 360, end = 360*2 + 180, scale_by = "curve_length")
spiral_track()
spiral_dendrogram(dend)
```

```{r, echo = FALSE, fig.width = 10, fig.height = 5}
dend = as.dendrogram(hclust(dist(runif(1000))))

p1 = grid.grabExpr({
    spiral_initialize(xlim = c(0, 1000), start = 360, end = 360*2 + 180, scale_by = "curve_length")
    spiral_track()
    spiral_dendrogram(dend)
})

p2 = grid.grabExpr({
    library(dendextend)
    dend = color_branches(dend, k = 4)
    spiral_initialize(xlim = c(0, 1000), start = 360, end = 360*2 + 180, scale_by = "curve_length")
    spiral_track()
    spiral_dendrogram(dend)
})
plot_grid(p1, p2)
```

`spiral_phylo()` works on the `phylo` object. In the following example, to cut the phylogenetic tree into several parts, 
the `phylo` object is converted to a `dendrogram` by the function `phylo_to_dendrogram()`, later `dendextend::cutree.dendrogram()`
is applied to obtain the splits.

```{r}
library(ape)
data(hivtree.newick)
tree.hiv = read.tree(textConnection(hivtree.newick))
n = length(tree.hiv$tip.label) # number of leaves

spiral_initialize(xlim =c(0, n), start = 360, end = 360*2 + 180, scale_by = "curve_length")
spiral_track()
spiral_phylo(tree.hiv)

split = cutree(phylo_to_dendrogram(tree.hiv), k = 8)
tb = table(split)
for(i in seq_along(tb)) {
    if(i == 1) {
        spiral_highlight(0, tb[1], gp = gpar(fill = i))
    } else {
        spiral_highlight(sum(tb[1:(i-1)]), sum(tb[1:i]), gp = gpar(fill = i))
    }
}
```

## Utility functions

There are several utility functions which help to get the information of current spiral plot. `spiral_info()` prints
the general information of current spiral:

```{r, echo = c(2, 3, 5)}
pdf(NULL)
spiral_initialize(xlim = c(0, 100))
spiral_track(ylim = c(-1, 1))
invisible(dev.off())
spiral_info()
```

The special variable `TRACK_META` retrieves several meta-information of the current track.

```{r}
TRACK_META
names(TRACK_META)
```

The following two functions convert the data coordinates to polar coordinates or the canvas coordinates
(the coordinates where the graphics are finally drawn).

- `xy_to_cartesian()`
- `xy_to_polar()`

The following three functions get or set the tracks:

- `current_track_index()`
- `set_current_track()`
- `n_track()`


## Session info

```{r}
sessionInfo()
```


<script src="jquery.min.js"></script>
<script src="jquery.sticky.js"></script>
<script>
$(document).ready(function(){
    $("#TOC").sticky({
        topSpacing: 0,
        zIndex:1000    
    })
    $("#TOC").on("sticky-start", function() {

        $("<p style='font-size:1.2em; padding-left:4px;'><a id='TOC-click'>Table of Content</a></p>").insertBefore($("#TOC ul:first-child"));
        $("#TOC-click").hover(function() {
            $(this).css("color", "#0033dd").css("cursor", "pointer");
            $("#TOC").children().first().next().show();
            $("#TOC").hover(function() {
                $(this).children().first().next().show();
            }, function() {
                $(this).children().first().next().hide();
                $("body").off("hover", "#TOC");
            })
        }, function() {
            $(this).css("color", "#0033dd");
        })
        $("#TOC").children().first().next().hide();

    })
    $("#TOC").on("sticky-end", function() {
        $("#TOC").children().first().remove();
        $("#TOC").children().first().show();
    })
});
</script>